Tsbrowse: an interactive browser for Ancestral Recombination Graphs

### 2025-07-04 manuscript/paper.tex

\journaltitle{Bioinformatics}
\journaltitle{Preprint}
\DOI{DOI HERE}
\copyrightyear{2025}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

\abstract{
    %\textbf{Motivation:}
    Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
    ancestry of a set of recombining sequences. The ability to capture the
    evolutionary history of samples makes ARGs valuable in a wide range of
    applications in population and statistical genetics. ARG-based approaches are
    increasingly becoming a part of genetic data analysis pipelines due to
    breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
    of visualisation tools, which are crucial for validating inferences and
    generating hypotheses.
    %\textbf{Results:} 
    We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
    visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
    and mutations. We demonstrate the application of \texttt{tsbrowse} to various
    data sources and scenarios, and highlight its key features of browsability
    along the genome, user interactivity, and scalability to very large sample
    sizes.\\ \textbf{Availability:}\\ Python package:
    \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
    \texttt{\url{https://github.com/tskit-dev/tsbrowse}}, \\ Documentation:
    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}, \\ DOI: 
    \texttt{\url{https://doi.org/10.5281/zenodo.15683039}}\\
    %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
    %\textbf{Supplementary information:} Available at
    %\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
    %online}.}}
}

@@

\subsection{Architecture}
\texttt{Tsbrowse} is a modular client-server application
written in Python, optimised for ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}).
The application has two basic commands: \texttt{preprocess}
and \texttt{serve}. The \texttt{preprocess} command
takes an input ARG file and augments the \texttt{tskit}
tables with additional columns,
precomputing all the information required for visualisation,
and storing it as a compressed \texttt{.tsbrowse} file.
%(The \texttt{.tsbrowse} file is also a valid input for
%\texttt{tszip}, a general utility for compressing \texttt{tskit}
%files.)
To visualise an ARG we then run \texttt{tsbrowse serve}, which
by default will open a browser window on the local machine, but
also supports running the server on a remote machine across the
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be
downloaded from the server.

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web browser
for display, and dynamically updates as the user interactively navigates the
ARG.
This approach allows us to summarise very large ARGs interactively;
%Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
Fig.~S1 for example, shows a screenshot of
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
    \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{Screenshot of the Edges view in \texttt{tsbrowse}.
        Visualisation of the edges in a simulated ARG with a strong selective
        sweep in the middle of the genome (see Supplement for details).
        Edges are shown in the main browser pane on the left, with additional
        histograms summarising the edges on the right.
        The effect of the sweep can be seen by the lack of edges crossing the
        centre of the simulated chromosome due to an excess of recent coalescent
        events. Moving away from the focal site, the oldest edges are the
        first to rejoin the ARG, followed by more recent edges, resulting in a
        wedge-like pattern of missing edges in the centre.
        \label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with views to
describe various aspects of the ARG.
Fig.~\ref{fig:Figure_2}
demonstrates the user interface using the Edges view as an example.
The plot on the left is an overview of the 33,929 edges
in a simulated ARG with a strong selective sweep (see
Supplement for details).
Each edge in the ARG is depicted as a horizontal line connecting the genomic coordinates
of the
parent and child nodes on the x-axis, and the y-axis shows the time of
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner provided by 
Bokeh~\citep{bokeh}, allowing them to pan and zoom as required. 
The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for
mutations
%(e.g. Fig~\ref{fig:Supplementary_Figure_1})
(e.g., Fig.~S1)
and nodes
%(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
(e.g., Fig.~S2).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
searching and sorting utilities, which is a valuable debugging utility for
developers.

@@

Interest in ARG-based methods is burgeoning, but the methods are new
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
%element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
element of the inference pipeline. Fig.~S4
shows a region of the 1000 Genomes data with gaps in site density that are spanned by
 exceptionally long edges, which are likely to bias downstream statistics.
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm.
%Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
Fig.~S2 shows the genomic spans of
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
excess of long haplotypes in the very ancient past. These insights
have helped guide development and
may lead to significant improvements in performance.

@@

\section*{Funding}
SK acknowledges full support and funding from Novo Nordisk Research Centre Oxford,
and funding from the Biotechnology and Biological Sciences Research Council 
(UKRI-BBSRC) [grant number BB/T008784/1]. DMR is funded by 
a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
JK acknowledges EPSRC (research grant EP/X024881/1),
NIH (research grants HG011395 and HG012473)
and the Robertson Foundation.

>>>>

\journaltitle{Bioinformatics}
%\journaltitle{Preprint}
\DOI{DOI HERE}
\copyrightyear{2025}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

%Summary, Availability and Implementation, Contact, and Supplementary information
\abstract{
    \textbf{Summary:}
    Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
    ancestry of a set of recombining sequences. The ability to capture the
    evolutionary history of samples makes ARGs valuable in a wide range of
    applications in population and statistical genetics. ARG-based approaches are
    increasingly becoming a part of genetic data analysis pipelines due to
    breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
    of visualisation tools, which are crucial for validating inferences and
    generating hypotheses. We present \texttt{tsbrowse}, an open-source,
    web-based Python application for the interactive visualisation of the fundamental 
    building blocks of ARGs, i.e., nodes, edges and mutations. We demonstrate the 
    application of \texttt{tsbrowse} to various data sources and scenarios, and highlight 
    its key features of browsability along the genome, user interactivity, and scalability 
    to very large sample sizes.\\\textbf{Availability and Implementation:}
    \texttt{Tsbrowse} is installed as a Python package from PyPI
    (\texttt{\url{https://pypi.org/project/tsbrowse/}}), while a development version 
    is maintained at \texttt{\url{https://github.com/tskit-dev/tsbrowse}}. Documentation 
    is available at \texttt{\url{https://tskit.dev/tsbrowse/docs/}}.
    Source code is archived on Zenodo with DOI, 
    \texttt{\url{https://doi.org/10.5281/zenodo.15683039}}\\
    \textbf{Contact:} \href{savita.karthikeyan@st-hughs.ox.ac.uk}{savita.karthikeyan@st-hughs.ox.ac.uk}\\
    \textbf{Supplementary information:} Available at
    \textit{Bioinformatics} online.
}

@@

\subsection{Architecture}
\texttt{Tsbrowse} is a modular application
written in Python, optimised for ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}).
The application has two basic commands: \texttt{preprocess}
and \texttt{serve}. The \texttt{preprocess} command
takes an input ARG file and augments the \texttt{tskit}
tables with additional columns,
precomputing all the information required for visualisation,
and storing it as a compressed \texttt{.tsbrowse} file.
%(The \texttt{.tsbrowse} file is also a valid input for
%\texttt{tszip}, a general utility for compressing \texttt{tskit}
%files.)
To visualise an ARG we then run \texttt{tsbrowse serve}, which
by default will open a browser window on the local machine, but
also supports running the server on a remote machine across the
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be
downloaded from the server.

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader~\citep{Datashader} and the wider
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web browser
for display, and dynamically updates as the user interactively navigates the
ARG.
This approach allows us to summarise very large ARGs interactively;
%Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
Fig.~S1 for example, shows a screenshot of
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
    \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{Screenshot of the Edges view in \texttt{tsbrowse}.
        Visualisation of the edges in a simulated ARG with a strong selective
        sweep in the middle of the genome (see Supplement for details).
        Edges are shown in the main browser pane on the left, with additional
        histograms summarising the edges on the right.
        The effect of the sweep can be seen by the lack of edges crossing the
        centre of the simulated chromosome due to an excess of recent coalescent
        events. Moving away from the focal site, the oldest edges are the
        first to rejoin the ARG, followed by more recent edges, resulting in a
        wedge-like pattern of missing edges in the centre.
        \label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with views to
describe various aspects of the ARG.
Fig.~\ref{fig:Figure_2}
demonstrates the user interface using the Edges view as an example.
The plot on the left is an overview of the 33,929 edges
in a simulated ARG with a strong selective sweep (see
Supplement for details).
Each edge in the ARG is depicted as a horizontal line connecting the genomic coordinates
of the
parent and child nodes on the x-axis, and the y-axis shows the time of
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner provided by 
%Bokeh~\citep{bokeh}, allowing them to pan and zoom as required. 
Bokeh (http://www.bokeh.pydata.org), allowing them to pan and zoom as required.
The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for
mutations
%(e.g. Fig~\ref{fig:Supplementary_Figure_1})
(e.g., Fig.~S1)
and nodes
%(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
(e.g., Fig.~S2).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
searching and sorting utilities, which is a valuable debugging utility for
developers.

@@

Interest in ARG-based methods is burgeoning, but the methods are new,
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the 
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
%element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
element of the inference pipeline. Fig.~S4
shows a region of the 1000 Genomes data with gaps in site density that are spanned by
 exceptionally long edges, which are likely to bias downstream statistics.
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm.
%Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
Fig.~S2 shows the genomic spans of
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
excess of long haplotypes in the very ancient past. These insights
have helped guide development and
may lead to significant improvements in performance.

@@

\section*{Funding}
SK acknowledges full support and funding from Novo Nordisk Research Centre Oxford
and the Biotechnology and Biological Sciences Research Council 
(UKRI-BBSRC) [grant number BB/T008784/1]. DMR is funded by 
a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
JK acknowledges EPSRC (research grant EP/X024881/1),
NIH (research grants HG011395 and HG012473)
and the Robertson Foundation.
